
```{r}
# Load the packages that is used in this project
#install.packages("car")
#install.packages("patchwork")
#install.packages("skimr")
#install.packages("broom")
#install.packages("tidyr")
#install.packages("pander")
#install.packages("stargazer")
library(stargazer)
library(broom)
library(skimr)
library(patchwork)
library(car)
library(ggplot2)
library(dplyr)
library(tidyr)
library(pander)
library(MASS)
library(gridExtra)
```


```{r}
Data = read.csv(file="CA_Weather_Fire_Dataset_1984-2025.csv", header=T, sep=",")
Data <- na.omit(Data)
Data$YEAR_adj <- Data$YEAR - min(Data$YEAR)


# 1. Convert Day of Year to Radians
# We use 2*pi because a full circle (year) is 2*pi radians.
# We divide by 365 (or 366 for leap years) to normalize it.
radians <- 2 * pi * Data$DAY_OF_YEAR / 365

# 2. Create the Sine Component
# Captures the "East-West" movement of the seasonal cycle
Data$Day_Sin <- sin(radians)

# 3. Create the Cosine Component
# Captures the "North-South" movement of the seasonal cycle
Data$Day_Cos <- cos(radians)

Data$FIRE_START_DAY <- as.factor(Data$FIRE_START_DAY)
CData <- Data[c(2:6,12,13,15:17)]

CData <- na.omit(CData)
```
### Estimating Trend/EDA

```{r echo=FALSE, fig.height=15, fig.width=20}
#EDA
histo_response <- ggplot(CData, aes(x = MAX_TEMP)) +
  geom_histogram(bins = 20, fill = "purple", color = "black", alpha = 0.7) +
  labs(
    title = "Distribution of Response Variable (MAX_TEMP)",
    x = "MAX_TEMP",
    y = "Degree in F"
  ) +
  theme_minimal()

box_response <- ggplot(CData, aes(y = MAX_TEMP)) +
  geom_boxplot(fill = "lightblue", color = "black", outlier.shape = 1) + # outlier.shape=1 shows outliers as circles
  labs(
    title = "Boxplot of Response Variable (MAX_TEMP)",
    y = "MAX_TEMP"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_blank(), # Remove x-axis text as there's only one box
        axis.ticks.x = element_blank()) # Remove x-axis ticks

(histo_response | box_response)
```


```{r echo=FALSE, fig.height=15, fig.width=20}
# --- 1. Create Individual Plots for Each Predictor ---

# Numerical Predictors (Histograms)
p1 <- ggplot(CData, aes(x = PRECIPITATION)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
  theme_minimal() +
  labs(title = "Precipitation", x = "Inches", y = "Count")

p2 <- ggplot(CData, aes(x = AVG_WIND_SPEED)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
  theme_minimal() +
  labs(title = "Avg Wind Speed", x = "Speed", y = "Count")

p3 <- ggplot(CData, aes(x = MIN_TEMP)) +
  geom_histogram(bins = 20, fill = "skyblue", color = "black") +
  theme_minimal() +
  labs(title = "Min Temp", x = "Temperature in F", y = "Count")

p4 <- ggplot(CData, aes(x = LAGGED_AVG_WIND_SPEED)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
  theme_minimal() +
  labs(title = "Lagged Avg Wind Speed in mph", x = "Speed", y = "Count")

p5 <- ggplot(CData, aes(x = LAGGED_PRECIPITATION)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
  theme_minimal() +
  labs(title = "Day of Year", x = "Day (1-365)", y = "Count")

p6 <- ggplot(CData, aes(x = YEAR_adj)) +
  geom_histogram(bins = 20, fill = "skyblue", color = "black") +
  theme_minimal() +
  labs(title = "Adjusted Year", x = "Year Index", y = "Count")

# Categorical Predictor (Bar Chart)
p7<- ggplot(CData, aes(x = FIRE_START_DAY)) +
  geom_bar(fill = "salmon", color = "black") +
  theme_minimal() +
  labs(title = "Fire Start Day", x = "Category", y = "Count")

# --- 2. Arrange Plots in a Grid ---
grid.arrange(p1, p2, p3, p4, p5, p6, p7, ncol = 3)
```

EDA ends

### Assumption Validating_Step 2(after checking for assumptions)

```{r}
# assumption validation methods
CData$MAX_TEMP_inv <- 1/CData$MAX_TEMP^1.2
CData$PRECIPITATION_trans <- 1/(CData$PRECIPITATION + 0.1)^3
CData$AVG_WIND_SPEED_trans <- log(CData$AVG_WIND_SPEED)
CData$LAGGED_PRECIPITATION_trans <- 1/(CData$LAGGED_PRECIPITATION + 0.1)^1.4
```

### Variable/Model Selection(after preliminary model was fitted and the assumption checks, which actually is just our final_model)

```{r}
final_model <- lm(MAX_TEMP_inv ~ Day_Sin + Day_Cos + MIN_TEMP + AVG_WIND_SPEED_trans + LAGGED_PRECIPITATION_trans + Day_Cos:MIN_TEMP + Day_Cos:AVG_WIND_SPEED_trans +
    LAGGED_AVG_WIND_SPEED + FIRE_START_DAY + PRECIPITATION_trans + AVG_WIND_SPEED_trans:PRECIPITATION_trans + LAGGED_PRECIPITATION_trans:PRECIPITATION_trans + MIN_TEMP:LAGGED_PRECIPITATION_trans +
    YEAR_adj, data = CData)
```

```{r}
BICdf <- CData[c(3, 5, 7:14)]
stepAIC(lm(MAX_TEMP_inv ~ Day_Sin + Day_Cos + MIN_TEMP + AVG_WIND_SPEED_trans + LAGGED_PRECIPITATION_trans + Day_Cos:MIN_TEMP + Day_Cos:AVG_WIND_SPEED_trans +
    LAGGED_AVG_WIND_SPEED + FIRE_START_DAY + PRECIPITATION_trans + AVG_WIND_SPEED_trans:PRECIPITATION_trans + LAGGED_PRECIPITATION_trans:PRECIPITATION_trans + MIN_TEMP:LAGGED_PRECIPITATION_trans +
    YEAR_adj, data = BICdf), direction = 'both', k = log(n))
```

```{r}
stepAIC(lm(MAX_TEMP_inv ~ Day_Sin + Day_Cos + MIN_TEMP + AVG_WIND_SPEED_trans + LAGGED_PRECIPITATION_trans + Day_Cos:MIN_TEMP + Day_Cos:AVG_WIND_SPEED_trans +
    LAGGED_AVG_WIND_SPEED + FIRE_START_DAY + PRECIPITATION_trans + AVG_WIND_SPEED_trans:PRECIPITATION_trans + LAGGED_PRECIPITATION_trans:PRECIPITATION_trans + MIN_TEMP:LAGGED_PRECIPITATION_trans +
    YEAR_adj, data = BICdf), direction = 'both', k = 2)
```

```{r}
aic_model <- lm(formula = MAX_TEMP_inv ~ Day_Sin + Day_Cos + MIN_TEMP + AVG_WIND_SPEED_trans + 
    LAGGED_PRECIPITATION_trans + Day_Cos:MIN_TEMP + Day_Cos:AVG_WIND_SPEED_trans + 
    LAGGED_AVG_WIND_SPEED + FIRE_START_DAY + PRECIPITATION_trans + 
    AVG_WIND_SPEED_trans:PRECIPITATION_trans + LAGGED_PRECIPITATION_trans:PRECIPITATION_trans + 
    MIN_TEMP:LAGGED_PRECIPITATION_trans + YEAR_adj, data = CData)
```


```{r}
bic_model <- lm(formula = MAX_TEMP_inv ~ Day_Sin + Day_Cos + MIN_TEMP + AVG_WIND_SPEED_trans + 
    LAGGED_PRECIPITATION_trans + LAGGED_AVG_WIND_SPEED + FIRE_START_DAY + 
    PRECIPITATION_trans + YEAR_adj + Day_Cos:AVG_WIND_SPEED_trans + 
    AVG_WIND_SPEED_trans:PRECIPITATION_trans + MIN_TEMP:LAGGED_PRECIPITATION_trans, 
    data = CData)
```


```{r}
AIC(bic_model)
BIC(bic_model)
AIC(final_model)
BIC(final_model)
# final_model is the model we manually choose, and the AIC automated selection agrees with us, and BIC does not. Since for both model AIC does not fully agree with BIC(only a small gap), and it turn out that our model has better coefficients of determination(R^2) and adjusted R^2 means that our model explains more variation than BIC selected model does, and also for BIC selected model the AIC does not agree with BIC too in about the same difference, our model is better at AIC, selected is better at BIC. In the consideration of meteorological knowledge in the real world, two removed interaction terms is useful and significantly affect each other(temperatures changes with different seasons, rains or not in the last 7 days does affect the prediction of precipitation today). Thus we decided to keep those two terms.
```

```{r}
# T Test and summary
# 1. Extract the coefficient table
coef_mat <- coef(summary(final_model))

# 2. Convert to Data Frame and fix Row Names
# We move the row names (predictors) into a real column named "Predictor"
coef_df <- as.data.frame(coef_mat)
coef_df <- tibble::rownames_to_column(coef_df, var = "Predictor")

# 3. Print with pander
pander(coef_df, 
       style = "rmarkdown", 
       split.table = Inf,       # Forces single table
       split.cells = Inf,       # Prevents wrapping text inside cells
       caption = "Summary of Final Model")
null_model <- lm(MAX_TEMP_inv ~ 1, data = CData)
# ANOVA Overall Test
anova(null_model, final_model)
# Partial F Test for bic autoselected model
anova(bic_model, final_model)
# Partial F test for lower p value predictors in the T test with its interaction terms
reduced_model <- lm(MAX_TEMP_inv ~ Day_Sin + Day_Cos + MIN_TEMP +  LAGGED_PRECIPITATION_trans + Day_Cos:MIN_TEMP +
    LAGGED_AVG_WIND_SPEED + FIRE_START_DAY + PRECIPITATION_trans  + LAGGED_PRECIPITATION_trans:PRECIPITATION_trans + MIN_TEMP:LAGGED_PRECIPITATION_trans +
    YEAR_adj, data = CData)
anova(reduced_model, final_model)
#The avrage wind speed has a large P value in T test, but it has interaction term with seasonality and precipitation, and their P value is extremely small, by the hierarchy principle, the effect of main effect term may shift to the interaction term, so avrage wind speed needs to stay in the model. The Partial F Test with a p value < 2.2e-16 also support my conclusion.
#ANOVA Overall Test with a p value < 2.2e-16 indicated that a statistically significant linear relationship exists for at least one predictor
#Partial F Test with a p value at 0.004976 and T test with p value at 0.02141, 0.01974 for Day_Cos:MIN_TEMP and LAGGED_PRECIPITATION_trans:PRECIPITATION_trans indicates that there exist a significant linear relationship between: both two interaction terms with response and two terms individually with response. Thus no predictor is needed to be delete from the model.
```

Variable/Model Selection ends
### Assumptions Check

```{r echo=FALSE, fig.height=50, fig.width=40}
# Plots parameters
standardized_r <- rstandard(final_model)
e_hat <- resid(final_model)
y_hat <- fitted(final_model)
residual_df <- data.frame(residuals = e_hat, std_resid = standardized_r)
residual_fitted_df <- data.frame(residuals = e_hat, fitted = y_hat, std_resid = standardized_r)
plot_df <- cbind(CData, residuals = e_hat, std_resid = standardized_r, DAY_OF_YEAR = Data$DAY_OF_YEAR)
pair_df <- cbind(CData, residuals = e_hat, std_resid = standardized_r)

# Define a border theme to separate plots visually ---
border_theme <- theme(
  plot.background = element_rect(color = "black", linewidth = 0.5, fill = NA)
)

# 1. The "Real" Check: Residuals vs. Day of Year
# This checks if your Sin/Cos terms successfully removed the seasonality
resid_vs_doy_plot <- ggplot(plot_df, aes(x = DAY_OF_YEAR, y = std_resid)) +
  geom_point(alpha = 0.3, color = "gray") +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_smooth(se = FALSE, color = "blue") + # Look for waves here
  labs(title = "Residuals vs. Day of Year",
       subtitle = "If the blue line is wavy, the Sin/Cos model missed some seasonality.",
       x = "Day of Year (1-365)", y = "Residuals") +
  theme_minimal()

# 2. Visualizing Interaction: Seasonality * Precipitation
# We categorize precipitation to see how the seasonal curve changes when it rains
interact_data <- plot_df %>%
  mutate(Rain_Status = ifelse(PRECIPITATION > 0, "Wet Day", "Dry Day"))

interact_season_plot <- ggplot(interact_data, aes(x = DAY_OF_YEAR, y = MAX_TEMP, color = Rain_Status)) +
  geom_point(alpha = 0.1) + 
  
  # Add the smooth trend lines for Wet vs Dry
  geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs"), se = FALSE, size = 1.5) +
  
  labs(title = "Interaction: How Rain Affects the Seasonal Cycle",
       subtitle = "Does the gap between Red (Dry) and Blue (Wet) change throughout the year?",
       x = "Day of Year", y = "Max Temperature (inversed form with inversed slope)") +
  theme_minimal()


# --- 1. Response vs Fitted (Observed vs Predicted) ---
# Check linearity and model accuracy
response_vs_fitted_plot <- ggplot(plot_df, aes(x = y_hat, y = MAX_TEMP_inv)) +
  geom_point(alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") + # The perfect prediction line
  labs(title = "Response vs Fitted",
       subtitle = "Ideally, points should hug the red dashed line",
       x = "Fitted Values (Predicted Max Temp)",
       y = "Observed Max Temp") +
  theme_minimal()

# --- 2. Normal Q-Q Plot ---
# Check normality of residuals
qq_plot <- ggplot(residual_df, aes(sample = std_resid)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "Normal Q-Q Plot of Residuals",
       subtitle = "Ideally, points should fall on the red line",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles") +
  theme_minimal()

# --- 3. Residuals vs Fitted ---
# Check for heteroscedasticity (fanning) and non-linearity
resid_vs_fitted_plot <- ggplot(residual_fitted_df, aes(x = fitted, y = std_resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residuals vs Fitted",
       subtitle = "Ideally, a random cloud centered around 0 (No patterns/cones)",
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal()

# --- 4. Residuals vs Each Predictor ---

# Numerical Predictor: Precipitation
resid_vs_precip_plot <- ggplot(plot_df, aes(x = PRECIPITATION_trans, y = std_resid)) +
  geom_point(alpha = 0.5) +
  labs(title = "Residuals vs Precipitation",
       x = "Precipitation", y = "Residuals") +
  theme_minimal()

# Numerical Predictor: Avg Wind Speed
resid_vs_wind_plot <- ggplot(plot_df, aes(x = AVG_WIND_SPEED_trans, y = std_resid)) +
  geom_point(alpha = 0.5) +
  labs(title = "Residuals vs Avg Wind Speed",
       x = "Avg Wind Speed", y = "Residuals") +
  theme_minimal()

# Numerical Predictor: Min Temp
resid_vs_temprange_plot <- ggplot(plot_df, aes(x = MIN_TEMP, y = std_resid)) +
  geom_point(alpha = 0.5) +
  labs(title = "Residuals vs Min Temp",
       x = "Min Temp", y = "Residuals") +
  theme_minimal()

# Numerical Predictor: Lagged Precipetation
resid_vs_ratio_plot <- ggplot(plot_df, aes(x = LAGGED_PRECIPITATION_trans, y = std_resid)) +
  geom_point(alpha = 0.5) +
  labs(title = "Residuals vs Lagged Precipetation",
       x = "Lagged Precipetation", y = "Residuals") +
  theme_minimal()

# Numerical Predictor: Lagged Wind Speed
resid_vs_lagwind_plot <- ggplot(plot_df, aes(x = LAGGED_AVG_WIND_SPEED, y = std_resid)) +
  geom_point(alpha = 0.5) +
  labs(title = "Residuals vs Lagged Avg Wind Speed",
       x = "Lagged Avg Wind Speed", y = "Residuals") +
  theme_minimal()

# Numerical Predictor: Day cos
resid_vs_day_cos_plot <- ggplot(plot_df, aes(x = Day_Cos, y = std_resid)) +
  geom_point(alpha = 0.5) +
  labs(title = "Residuals vs Day cos transformation",
       x = "Day cos", y = "Residuals") +
  theme_minimal()

# Numerical Predictor: Day sin
resid_vs_day_sin_plot <- ggplot(plot_df, aes(x = Day_Sin, y = std_resid)) +
  geom_point(alpha = 0.5) +
  labs(title = "Residuals vs Day sin transformation",
       x = "Day sin", y = "Residuals") +
  theme_minimal()

# Numerical Predictor: Adjusted Year
resid_vs_year_plot <- ggplot(plot_df, aes(x = YEAR_adj, y = std_resid)) +
  geom_point(alpha = 0.5) +
  labs(title = "Residuals vs Year (Adj)",
       x = "Year (Adjusted)", y = "Residuals") +
  theme_minimal()

# Categorical Predictor: Fire Start Day (Boxplot)
resid_vs_fire_plot <- ggplot(plot_df, aes(x = FIRE_START_DAY, y = std_resid)) +
  geom_boxplot(aes(fill = FIRE_START_DAY)) +
  labs(title = "Residuals vs Fire Start Day",
       x = "Fire Start Day", y = "Residuals") +
  theme_minimal()

plot_list <- list(
  response_vs_fitted_plot, qq_plot, resid_vs_fitted_plot,
  resid_vs_doy_plot, interact_season_plot, resid_vs_day_sin_plot, resid_vs_day_cos_plot,
  resid_vs_temprange_plot, resid_vs_precip_plot, resid_vs_ratio_plot,
  resid_vs_wind_plot, resid_vs_lagwind_plot, resid_vs_year_plot, resid_vs_fire_plot
)

# You can skip this loop if you don't want borders, but it helps alignment
plots_with_borders <- lapply(plot_list, function(p) p + border_theme)

# --- Construct the Patchwork Layout ---

# Row 1: Model Performance (3 Plots)
row1 <- plots_with_borders[[1]] | plots_with_borders[[2]] | plots_with_borders[[3]]

# Row 2: Seasonality & Cyclical Terms (4 Plots - Tightly Grouped)
# Day of Year | Interaction | Sine | Cosine
row2 <- plots_with_borders[[4]] | plots_with_borders[[5]] | plots_with_borders[[6]] | plots_with_borders[[7]]

# Row 3: Temperature & Precipitation (3 Plots)
# Min Temp | Precip | Lagged Precip
row3 <- plots_with_borders[[8]] | plots_with_borders[[9]] | plots_with_borders[[10]]

# Row 4: Wind & Other Controls (4 Plots)
# Wind | Lagged Wind | Year | Fire
row4 <- plots_with_borders[[11]] | plots_with_borders[[12]] | plots_with_borders[[13]] | plots_with_borders[[14]]

# --- Combine All Rows ---
final_dashboard <- row1 / row2 / row3 / row4

# Add a main title and display
final_dashboard + 
  plot_annotation(
    title = 'Diagnostic Dashboard: Max Temp Regression Model',
    subtitle = 'Row 1: Model Fit | Row 2: Seasonality (Sin/Cos) | Rows 3-4: Physical Predictors',
    theme = theme(plot.title = element_text(size = 20, face = "bold"))
  )
```

```{r}
pairs(plot_df[c(3,5,7:10,12:14)])
```
Assumptions Checking ends
### Assumptions Validating_Step 1

```{r}
# based on "correcting the worst violation first" from lecture, we fix the normality first by apply boxcox transformation on response
pre_model <- lm(MAX_TEMP ~ Day_Sin + Day_Cos + MIN_TEMP + AVG_WIND_SPEED + LAGGED_PRECIPITATION + Day_Cos:MIN_TEMP + Day_Cos:AVG_WIND_SPEED +
    LAGGED_AVG_WIND_SPEED + FIRE_START_DAY + PRECIPITATION + AVG_WIND_SPEED:PRECIPITATION + LAGGED_PRECIPITATION:PRECIPITATION + MIN_TEMP:LAGGED_PRECIPITATION +
    YEAR_adj, data = CData)
boxcox(pre_model)
# Then fixing the non constant variance violation on predictors using powerTransFormation, using bcnPower for predictors containing 0
powerTransform(CData$PRECIPITATION, family = 'bcnPower')
powerTransform(CData$LAGGED_PRECIPITATION, family = 'bcnPower')
powerTransform(CData$AVG_WIND_SPEED, family = 'bcPower')
```

Assumptions Validating ends

### Diagnostics

```{r}
# multicollinearity, look at the third term (adjusted gvif) if its greater than 2.5, the outputs suggests no muticollinearity presents, the value beyond the cufoff is called structural multicollinearity
# and its mathematical harmless
vif(final_model, type = 'predictor')
```
```{r}
# assessing problematic observations
p <- length(coef(final_model)) - 1
n <- nrow(CData)
# leverages cutoff
hii_cutoff <- 2 * (p + 1)/n
# standardized residuals cutoff is [-4, 4]
# Cook's difference cutoff
Di_cutoff <- qf(0.5, p + 1, n - p - 1)
#DFFITS cutoff
dffits_cutoff <- 2 * sqrt((p + 1)/n)
#DFBETAS cutoff
dfbetas_cutoff <- 2/sqrt(n)
```


```{r}
# assessing problematic observations
hii <- hatvalues(final_model)
ri <- rstandard(final_model)
Di <- cooks.distance(final_model)
dffits <- dffits(final_model)
dfbetas <- dfbetas(final_model)
# leverages
leverages <- which(hii > hii_cutoff)
# outliers
outliers <- which(ri < -4 | ri > 4)
# influence points:
# Cook's distance 
affect_all_fitted <- which(Di > Di_cutoff)
# DFFITS
affect_individual_fitted <- which(abs(dffits) > dffits_cutoff)
# DFBETAS

dfb_long <- as.data.frame(as.table(dfbetas))
colnames(dfb_long) <- c("Row_Index", "Predictor", "DFBETA_Value")

# 1. Filter for Violations
# We only keep rows where the absolute value exceeds the cutoff
violations <- dfb_long[abs(dfb_long$DFBETA_Value) > dfbetas_cutoff, ]

# 2. Sort by Severity (Descending)
violations <- violations[order(abs(violations$DFBETA_Value), decreasing = TRUE), ]

# 3. View the Top Offenders
head(violations, 20)
# Since there's no strong enough evidence to remove the problematic observations (no invalid observations, no extreme values, the range of all observations is consistent with the real world situation). These observations will be kept.
```
Diagonostics Ends

### Model Goodness


```{r}
# 1. Calculate AIC and BIC manually
aic_val <- AIC(my_model)
bic_val <- BIC(my_model)
```



```{r}
# Model validation and goodness measurements via LOOCV
Y <- CData$MAX_TEMP_inv
Y_hat <- final_model$fitted.values
CV_n <- mean(((Y-Y_hat)/(1-hii))^2)
CV_n
# 2.36582e-07, close to the MSE of our model, thus our model is validated.
```
Model Goodness ends

